The unphysicality of Hilbert spaces

### 2023-06-21 Articles/2023-HilbertQuantum/HilbertQuantum.tex
Reviewed main text

If we require, then, that the theory reproduces the properties of quantum mixtures, how they can be decomposed in terms of pure states and what their entropy is, the complex inner product space is recovered as a structure that recovers the correct properties. As we said, a physical theory must tell us how to properly characterized statistical mixture, therefore the requirement is necessitated by the physics. \textbf{The inner product space structure is physical}.

@@

Completeness of a Hilbert space tells us that the space includes the limit of every Cauchy sequence. This property is more difficult to characterize physically. We can look at an equivalent characterization:\footnote{Roman 2008, p. 330 Theorem 13.8

	Roman, Stephen (2008), Advanced Linear Algebra, Graduate Texts in Mathematics (Third ed.), Springer, ISBN 978-0-387-72828-5} the space is complete if, given a series of vectors
$$ \sum _{k=0}^{\infty }u_{k}$$
such that
$$ \sum _{k=0}^{\infty }\|u_{k}\|<\infty \,$$
the series converges in the space. This is now in terms of linear combinations, and therefore it is essentially telling us that we can perform a linear superposition if the sum of the norm of the coefficient converge. We are basically extending superposition to infinite sums. Is this a physical requirement of just mathematical convenience?

As we saw before, superposition experimentally means splitting the state along multiple paths. The above mathematical expression means splitting the original state into infinitely many progressively small component, and be able to place them where we please. It also assumes, for example, that we can place them over an infinite spatial range. Since the limit must be in the space, we are literally saying that we can prepare states that are spread over infinite regions. This does not sound physical.

Again, given the difficulties of understanding superposition, let us reframe the result in terms of statistical quantities. Suppose that we are able to prepare pure states with arbitrary finite expectation value for position. Then completeness means we are also able to prepare pure states that falls off at infinity as $\frac{1}{x^2}$. The expectation value of $x$ would the integral of a function that falls off as $\frac{1}{x}$ which diverges. Clearly, this does not make physical sense. When we posit that a quantity, like position, has an infinite range of possible values, we simply mean that observed value can be arbitrarily large, not that we can literally spread values over an infinite range.

The argument, then, goes like this. On physical grounds, given a quantity that can take arbitrarily large values, like position, for each possible value there must exist some state whose expectation matches that value. On mathematical grounds, we require completeness, which forces us to posit states with infinite expectation value. But this, on physical grounds, is not teneable. Therefore \textbf{completeness in unphysical.}

@@

We are forced to conclude that Hilbert spaces have unphysical requirements when applied to quantum mechanics. We clearly see that the issue is the closure of operation over infinite sequences, which is not surprising.\footnote{Large branches of mathematics were developed to handle infinities, both large and small. Weyl stated ``Mathematics is the science of the infinite, its goal the symbolic comprehension of the infinite with human, that is finite, means.'' Hermann Weyl, "The Open World: Three Lectures on the Metaphysical Implications of Science," (1932) as quoted in Mind and Nature: Selected Writings on Philosophy, Mathematics, and Physics (2009) ed. Peter Pesic.} Now we have to understand how big of a problem it is.

On physical grounds, the state space of quantum system is the projective space of a complex inner product space. If the space is finite dimensional, the space is automatically complete.\footnote{This is exactly why, mathematically, completeness is nice as it makes infinite dimensional space behave like finite dimensional ones.} If not, we can always complete it by enlarging the space. In general, then, it is true that a quantum state is a ray of a Hilbert space, but the converse is not true: not all rays of the same Hilbert space will be valid quantum states.

This may not seem like a problem: we simply have a bigger set of object of what we strictly need. Unfortunately, in mathematics working on a bigger space is not always a good thing. For example, take the problem of defining volumes in three dimensional euclidean space $\mathbb{R}^3$. Formally, we want to find a measure $\mu(U)$ that returns the size of the region $U$. Naively, we may want to define it on the set of all possible subsets. In this case, Banach–Tarski paradox tells us that we can take a unit ball, cut it into five pieces, and reassemble those pieces into two balls of the same size. Clearly, this is a problem. However, if we restrict ourselves with a measure on the Borel sets instead of all sets, we get a consistent definition. Therefore, in mathematics, bigger is not always better.\footnote{In another work, we show that the Borel sets are exactly those sets that can be associated with experimental procedures. In many cases, we have found that physical requirements naturally map to well behaved spaces.}

The above example is just a cautionary tale and it doesn't tell us what happens in this specific case. Therefore, let us give an example of the things that can go wrong. Unitary transformation play an important role in the theory of Hilbert spaces for at least two reasons. The first is that two Hilbert spaces that differ by a unitary transformation are mathematically equivalent. A change of representation, like a change of basis, is a unitary transformation. The second reason is that time evolution for an isolated system is represented by a unitary transformation. Let us see, then, the interplay between unitary transformations and our previous findings.

In a Hilbert space, we can always find a unitary transformation that maps one vector to any other. Therefore, there will be unitary transformations that map states with finite expectation values to infinite ones. This means that whether something has a finite expectation value or not will depend on the choice of representation. As a concrete example, consider the following two wave functions
\begin{align}
	\psi(x) &= \sqrt{\frac{e^{-x^2}}{\sqrt{\pi}}} \\
	\phi(x) &= \frac{1}{\sqrt{\pi(x^2 + 1)}}.
\end{align}
Both states are symmetric, and therefore will have zero expectation value for $x$. The first distribution is a Gaussian, and therefore all moments of the distribution, the expectation for $x^n$, are finite. The second goes to infinity like $\frac{1}{x^2}$, so the expectation will diverge for all. We can now find a change of variable $y(x)$ that transforms one distribution into the other. We need to set
\begin{equation}
	\begin{aligned}
	\int_{0}^{x} \psi^\dagger(x) \psi(x) dx &= \int_{0}^{y(x)} \phi^\dagger(x) \phi(x) dx \\
	\int_{0}^{x} \frac{e^{-x^2}}{\sqrt{\pi}} dx &= \int_{0}^{y(x)} \frac{1}{\pi(y^2 + 1)} dx \\
	\frac{\erf(x)}{2} &= \frac{\tan^{-1}(y)}{\pi} \\
	y &= \tan \left(\frac{\pi}{2}\erf(x)\right). \\
	\end{aligned}
\end{equation}
Therefore, thorough a variable change, we can change a state that has finite expectation value for all moments greater than the first into a state that has infinite expectation value for the same moments. Variable changes, in terms of the Hilbert space, are unitary transformation. Therefore the theory of Hilbert spaces, as applied to quantum mechanics, would tell us that the two representations are equivalent.

Experimentally, these are not equivalent. Identifying a state through expectation value is a common technique, called quantum tomography. If a unitary transformation can change finite expectations to infinite ones, we are effectively changing what can be distinguished through a tomography. Not only we cannot measure infinite values, but different finite values would get lumped into the same infinite value. In the above example, if we increase the variance, all even moments would change, yet, after the transformation, they would all be infinite. Therefore the idea that such unitary transformation links to physically equivalent representations is, on physics ground, untenable.

Moreover, we can create the following family of transformations
\begin{equation}
	y = \cos(\omega t) x + \sin(\omega t) \tan \left(\frac{\pi}{2}\erf(x)\right).
\end{equation}
This is clearly continuous in $t$ and, for $t=0$, we have the identity transformation and therefore, mathematically, there is no reason why this wouldn't be a valid time evolution. The effect is simply a stretching back and forth the wavefunctions, such that the state keeps oscillating between the above two functions. On physics grounds, however, we are going from finite to infinite expectations in finite time. This, again, is physically untenable.

We hope that this gives an idea of the type of issues embedded in the mathematical definitions. When performing actual calculations, we do not see these issues because the physics that we are describing makes us work in a corner of the structure where things are reasonable. But when one tries to prove theorems or general results on quantum theory, one currently is forced to consider the entire Hilbert space. It is not inconceivable to expect that results that would be reasonable physically fail to valid when enlarging the space of possible states to those that have infinite expectation values and the space of physical transformations to those that map finite to infinite expectations. We therefore wonder whether some of the issues we have in quantum field theory (e.g. the inability to find a valid measure for path integrals, inequivalent representation of interacting theories (?), infinities) are simply a consequence of the use, at a fundamental level, of unphysical objects.

@@

Suppose the reader agrees with our finding, the subsequent question is: what is the alternative? If we are only interested in observable with finitely many possible outcomes, we can use finite dimensional spaces. But as soon we have discrete quantities with arbitrarily large values, such as the number of particles, or quantities that are continuous, like position and momentum, we need infinite dimensional spaces. If we allow infinite dimensional spaces, we need some way of controlling what infinite linear combinations are allowed or not allowed. Cauchy completeness is one such way, but if that is not physical, what should we use?

@@

@@

>>>>

The argument can be summed up as follows. On physical grounds, for any physical theory, we must be able to prepare statistical mixtures and characterize their entropy. The complex inner product space structure is exactly required to model the correct relationships of quantum mixtures, therefore the mathematical definition is necessitated by physical requirements. \textbf{The inner product space structure is physical}.

@@

We now turn to the last property of Hilbert spaces: completeness. Mathematically, this means that the space includes the limit of every Cauchy sequence; physically, this property is more difficult to characterize. We can look at an equivalent characterization:\footnote{Roman 2008, p. 330 Theorem 13.8

	Roman, Stephen (2008), Advanced Linear Algebra, Graduate Texts in Mathematics (Third ed.), Springer, ISBN 978-0-387-72828-5} the space is complete if, given a series of vectors
$$ \sum _{k=0}^{\infty }u_{k}$$
such that
$$ \sum _{k=0}^{\infty }\|u_{k}\|<\infty \,$$
the series converges in the space. This is now in terms of linear combinations, and therefore it is essentially telling us that we can prepare states that are superposition of infinitely many elements or, equivalently as we have seen, mixed states comprised of infinitely many pure states. Is this a physical requirement of just mathematical convenience?

Experimentally, we can imagining creating a superposition by splitting a beam along multiple paths, which is something typical, for example, in quantum optics.\footnote{TODO: cite} The above mathematical expression means splitting the original state into infinitely many progressively small components, and be able to place them where we please. However, it also assumes that we can place them over an infinite spatial range: we are literally saying that we can prepare states that are spread over infinite regions. This does not sound physical.

Again, given the difficulties of understanding superposition, let us reframe the result in terms of statistical quantities. Suppose that we are able to prepare pure states with arbitrary finite expectation value for position. Then completeness means we are also able to prepare pure states that falls off at infinity as $\frac{1}{x^2}$. The expectation value of $x$ would the integral of a function that falls off as $\frac{1}{x}$ which diverges. Clearly, this does not make physical sense. When we posit that a quantity, like position, has an infinite range of possible values, we simply mean that observed value can be arbitrarily large, not that we can literally spread values over an infinite range.\footnote{This is the difference between potential infinity and actual infinity.}

The argument can be summed up as follows. On physical grounds, we assume the existence of quantities, like position, that can take arbitrarily large values. This requires that for each possible value there must exist some state whose expectation matches that value. Requiring completeness mathematically forces us to posit states with an infinite expectation value, which cannot be implemented physically. The mathematical definition implicitly assumes objects that cannot be physically realized. \textbf{The use of completeness for the space of quantum systems in unphysical.}

@@

It should not be surprising that closure over infinite sequences is problematic, as properly handling infinity is one of the key problems in mathematics.\footnote{Weyl stated ``Mathematics is the science of the infinite, its goal the symbolic comprehension of the infinite with human, that is finite, means.'' Hermann Weyl, "The Open World: Three Lectures on the Metaphysical Implications of Science," (1932) as quoted in Mind and Nature: Selected Writings on Philosophy, Mathematics, and Physics (2009) ed. Peter Pesic. Several branches of mathematics were developed to handle infinities, both large and small. } We should understand the impact on the physics of using completeness for that goal.

If an inner product space is finite dimensional, it is automatically complete.\footnote{This is exactly why, mathematically, completeness is nice as it makes infinite dimensional space behave like finite dimensional ones.} If the space is not complete, we can always complete it by enlarging the space. The prevalent attitude among physicists is that this pose no problem: we simply have a bigger set of objects than what we strictly need, and we simply discard what we do not need. Unfortunately, having a bigger space often creates bigger problems.

For example, take the problem of defining volumes in three dimensional euclidean space $\mathbb{R}^3$. Formally, we want to find a measure $\mu(U)$ that returns the size of the region $U$. Naively, we may want to define it on the set of all possible subsets. In this case, the Banach–Tarski paradox tells us that we can take a unit ball, cut it into five pieces, and reassemble those pieces into two balls of the same size. Clearly, this does not make physical sense. However, if we restrict ourselves with a measure on the Borel sets instead of all sets, we get a consistent definition. Therefore, in mathematics, bigger is not always better:\footnote{In another work, we show that the Borel sets are exactly those sets that can be associated with experimental procedures. In many cases, we have found that physical requirements naturally map to well behaved spaces.} we need to look more closely at what the math dragged in.

Unitary transformation play an important role in the theory of Hilbert spaces for at least two reasons. The first is that two Hilbert spaces that differ by a unitary transformation are mathematically equivalent. A change of representation, like a change of basis, is a unitary transformation. The second reason is that time evolution for an isolated system is represented by a unitary transformation. Let us see, then, the interplay between unitary transformations and our previous findings.

In a Hilbert space, we can always find a unitary transformation that maps one vector to any other. Therefore, there will be unitary transformations that map states with finite expectation values to infinite ones. This means that whether something has a finite expectation value or not will depend on the choice of representation. As a concrete example, consider the following two wave functions
\begin{align}
	\psi(x) &= \sqrt{\frac{e^{-x^2}}{\sqrt{\pi}}} \\
	\phi(x) &= \frac{1}{\sqrt{\pi(x^2 + 1)}}.
\end{align}
Both states are symmetric, and therefore will have zero expectation value for $x$. The first distribution is a Gaussian, and therefore all moments of the distribution, the expectation for $x^n$, are finite. The second goes to infinity like $\frac{1}{x^2}$, so the expectation will diverge for all moments above the first. We can now find a change of variable $y(x)$ that transforms one distribution into the other. We need to set
\begin{equation}
	\begin{aligned}
	\int_{0}^{x} \psi^\dagger(x) \psi(x) dx &= \int_{0}^{y(x)} \phi^\dagger(x) \phi(x) dx \\
	\int_{0}^{x} \frac{e^{-x^2}}{\sqrt{\pi}} dx &= \int_{0}^{y(x)} \frac{1}{\pi(y^2 + 1)} dx \\
	\frac{\erf(x)}{2} &= \frac{\tan^{-1}(y)}{\pi} \\
	y &= \tan \left(\frac{\pi}{2}\erf(x)\right). \\
	\end{aligned}
\end{equation}
Therefore, through a variable change, we can change a state that has finite expectation value for all moments greater than the first into a state that has infinite expectation value for the same moments. Variable changes, in terms of the Hilbert space, are unitary transformation. Therefore the theory of Hilbert spaces, as applied to quantum mechanics, would tell us that the two representations are equivalent.

On physical grounds, these are clearly not equivalent. Identifying a state through expectation value is a common technique (i.e quantum tomography). If a unitary transformation can change finite expectations to infinite ones, we are effectively changing what can be distinguished experimentally. Not only we cannot measure infinite values, but different finite values would get lumped into the same infinite value. In the above example, if we increased the variance, one observer will see all the even moments change, yet the other would see them all equally infinite. The math, as it stands, lumps together physically inequivalent representations into the same equivalence class, making it impossible later to separate physical problem from unphysical ones.

Moreover, we can create the following family of transformations
\begin{equation}
	y(t) = \cos(\omega t) x + \sin(\omega t) \tan \left(\frac{\pi}{2}\erf(x)\right).
\end{equation}
This is clearly continuous in $t$ and, for $t=0$, we have the identity transformation and therefore, mathematically, there is no reason why this wouldn't be a valid time evolution. What this does is stretching back and forth the wavefunction, such that the state keeps oscillating between the above two functions, making the expectation value oscillate from finite to infinite in finite time. This, again, is physically untenable.

Again, most of the time these issues do not arise when performing actual calculations because the physics makes us work in a corner of the structure where things are reasonable. The issue is that when one tries to prove theorems or general results on quantum theory, one currently is forced to consider the entire Hilbert space. It is unreasonable to expect that all theorems valid on all physical cases would have valid generalizations on unphysical ones, and therefore we are likely missing useful results. We therefore wonder whether many of the technical issues we have in quantum theory (e.g. non existence of a valid measure for path integrals, inequivalent representation of interacting theories, non existence of the categorial tensor product for Hilbert spaces, having to deal with infinities and infinite correlations)\footnote{TODO: cite} are simply a consequence of the use, at a fundamental level, of unphysical objects.

@@

Suppose the reader agrees with our finding, the subsequent question is: what is the alternative? We do need infinite dimensional spaces to handle unbound quantities, such as number of particles, or continuous quantities, such as position. If we can't use Cauchy completeness, how do we characterize the converge of infinite sequences?

A solution could be simply focusing on the expectation values of observables, and requiring a sufficient number of them to be finite. For example, we could require the expectation of all polynomial of position and momentum to be finite. In quantum mechanics, this means requiring all expectations of $\frac{1}{2}(X^nP^m - P^mX^n)$ to be finite. This restricts our space to those functions that are infinitely smooth (i.e. all derivative exist) and decrease very rapidly as $x$ goes to infinity: these are the Schwartz functions. Is the Schwartz space a better candidate?

First of all, the Schwartz space is a subspace of the $L^2$ Hilbert space taken as an inner product space. That is, all Schwartz functions are elements of the Hilbert space, they are closed under linear combinations and have the same inner product. It is not, however, complete, which is what we want. Moreover, it is a dense subspace, meaning that any element of the Hilbert space can always be approximated by a Schwartz function with an arbitrary level of precision. This is a good hint that we have thrown away only unphysical elements.

If we restrict ourselves to the Schwartz space, then, the only unitary transformations we are interested in are those that map Schwartz function to Schwartz functions. The change of coordinate we defined in the previous section, then, is ruled out. Limits in terms of Cauchy sequences are not guaranteed, but limits in terms of expectation values are guaranteed: if we have a sequence of states such that the limit of the expectation converge, then we are guaranteed that the limit is also a Schwartz function. Schwartz functions, in fact, are uniquely identified by the expectation of all polynomial of position and momentum, effectively guaranteed that they can be distinguished by quantum tomography.

The Schwartz space has also a very important property: it is the only space that is closed under the Fourier transform. That is, the Fourier transform of a Schwartz function is a Schwartz function, and it is the only set of functions with this property. The theory of tempered distribution, which allows us to define the Dirac $\delta$-function, is mathematically based, at its core, over the Schwartz space. Therefore, this space already plays a fundamental role in the definitions of the tools we already use.

While determining whether Schwartz spaces are the correct space to capture quantum states goes beyond the scope of this article, we do want to show that alternatives exist. Exploring these alternatives fully, to see exactly what theorems still hold which do not, and whether others are possible, is a significant effort that is extremely important and long overdue.

@@

We have made the case that Hilbert spaces in quantum mechanics are \textbf{unphysical} as they necessarily bring in states with infinite expectation values, coordinate transformations that turn finite expectations in infinite ones, and unitary transformations that do the same in finite time. Theorems that are valid on Hilbert spaces must be valid over these unphysical cases, meaning that we are possibly missing results and techniques that would be physically interesting.

The more general problem is that the physics community has become complacent in simply accepting mathematical definitions, without properly understanding their limit of applicability for physical theories. We believe that meaningful progress on the foundations of physics cannot happen without understanding the implicit assumptions embedded in the most basic mathematical tools.

@@

\section{TODO}

of a We saw that the main problem is having expectation values that diverge,

If we are only interested in observable with finitely many possible outcomes, we can use finite dimensional spaces. But as soon we have discrete quantities with arbitrarily large values, such as the number of particles, or quantities that are continuous, like position and momentum, we need infinite dimensional spaces. If we allow infinite dimensional spaces, we need some way of controlling what infinite linear combinations are allowed or not allowed. Cauchy completeness is one such way, but if that is not physical, what should we use?
